Fully structured proof theory for intuitionistic modal logics

### 2020-06-02 BicolorNS/aiml20/aiml20.tex
complete draft with Sec 5

This lead us to develop a \emph{fully structured} approach to intuitionistic modal proof theory capturing both the modal accesibility relation and the intuitionistic pre-order relation. 
%
A fully labelled framework, described succintly in Section~\ref{sec:labelled}, has already allowed us to obtain modular systems for all intuitionistic Scott-Lemmon logics~\cite{marin:morales:strassburger:hal}. 
%
In an attempt to make this system amenable for proof-search and decision procedures, we have started investigated a fully nested framework, presented in Section~\ref{sec:nested}.
%
We would be particularly interested in a suitable system for logic $\mathsf{IS4}$ whose decidability is not known.

%We present a labelled sequent system and a nested sequent system for intuitionistic modal logics capturing both the modal accesibility relation and the intuitionistic pre-order relation. 
%
%In the labelled calculus we made explicity both relations with $\rel$ and $\le$ into the system. 
%
%In the nested sequent system, we represent this relations with one bracket in a sequent (for the accesibility relation) and a with two brackets in a sequent (preorder relation).  
%

@@

We write $\lb x \Vdash \fm a$ if $\fm a \in V(\lb x)$ and, by definition, it is never the case that $\lb x \Vdash \fm{\BOT}$. %and always have $\lb w \Vdash \fm{\TOP}$
%
The relation $\Vdash$ is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:\vspace*{-.1cm}
\begin{align}
	&\lb x \Vdash \fm{A \AND B} && \text{iff}\qquad 
	\lb x \Vdash \fm A \text{ and } \lb x \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{A \OR B} && \text{iff}\qquad 
	\lb x \Vdash \fm A \text{ or } \lb x \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{A \IMP B} && \text{iff}\qquad 
	\text{for all } \lb y \text{ with } \futs xy\text{, if }\lb y \Vdash \fm A\text{ then }\lb y \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{\BOX A} && \text{iff}\qquad
	\text{for all }\lb y\text{ and }\lb z\text{ with }\futs xy\text{ and }\accs yz, \lb z \Vdash \fm A %\hfill $(\ast)$
	\tag{$\star$}\label{eq:box}\\
	&\lb x \Vdash \fm{\DIA A} && \text{iff}\qquad
	\text{there exists a }\lb y\text{ such that }\accs xy\text{ and }\lb y \Vdash \fm A
	\nonumber
\end{align}

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
		%		
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%
		\item\label{iii} $A$ is provable in $\labIKp$
		%
		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}

@@

\begin{definition}
	A two-sided intuitionistic \emph{fully nested sequent} is constructed from the grammar:
	$$\Cx \coloncolonequals \varnothing \mid \lf A, \Cx \mid \rt A, \Cx \mid \BR{\Cx} \mid \bBR{\Cx}$$
	where the $\lf{(\cdot)}$ and $\rt{(\cdot)}$ annotations are used to indicate that the formulas would occur on the left-hand-side or right-hand-side of a sequent respectively in the absence of the sequent arrow. 
\end{definition}

@@

A great interest of a nested system is that the rules for axiomatic extensions are usually more suitable for proof-search procedures that in the labelled setting.
%
Previous nested systems for intuitionistic modal logics~\cite{strassburger2013,kuznets:strassburger:maehara} can be extended from $\IK$ to $\mathsf{IS4}$ by simply adding the following rules:
\[\small
\vlinf{\lrn[t]\BOX}{}{\Cx{\lf{\BOX A}}}{\Cx{\lf{\BOX A}, \lf A}}
\quad
%\vlinf{\lrn[2]\BOX}{}{\Cx[1]{\lf{\BOX A}, \bBR{\BR{\Cx[2]}}}}{\Cx[1]{\bBR{\BR{\lf{\BOX A}, \Cx[2]}}}}
\vlinf{\lrn[4]\BOX}{}{\Cx[1]{\lf{\BOX A}, \BR{\Cx[2]}}}{\Cx[1]{\lf{\BOX A}, \BR{\lf{\BOX A}, \Cx[2]}}}
\quad
\vlinf{\rrn[t]\DIA}{}{\Cx{\rt{\DIA A}}}{\Cx{\rt{\DIA A}, \rt A}}
\quad
\vlinf{\rrn[4]\DIA}{}{\Cx[1]{\rt{\DIA A}, \BR{\Cx[2]}}}{\Cx[1]{\rt{\DIA A}, \BR{\rt{\DIA A}, \Cx[2]}}}
\]

@@

%\section{Extensions}
%
%\paragraph{Labelled}
%
%the frame
%condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
%inference rule
%\begin{equation}
%\label{eq:Rtrans}
%\vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
%	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
%\end{equation}
%where $\B$ stands for a set of relational atoms, and $\Gamma$ and
%$\Delta$ for multi-sets of labelled formulas.
%
%The transitivity rule in~\eqref{eq:Rtrans} can be axiomatised by the conjunction of the two
%versions of the $\ax 4$-axiom
%\begin{equation}
%\label{eq:4ax}
%\fm{\BOX A\IMP \BOX\BOX A}
%\qquad
%\fm{\DIA\DIA A\IMP\DIA A}
%\quad
%\end{equation}
%which are equivalent in classical
%modal logic. However, in intuitionistic modal logic they are not
%equivalent, and even though the logic $\IK\ax 4$, i.e., the
%intuitionistic version of the modal logic $\K\ax 4$, contains both
%axioms, they can also be added independently to the logic $\IK$ (an
%intuitionistic version of $\K$). The proof theory of these
%new logics has not been studied before; no existing labelled (or
%label-free) proof system can handle them, even though the
%corresponding frame conditions
%\begin{equation}
%\label{eq:4frame}
%\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
%\quad
%\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
%\;,
%\end{equation}
%%  \todo{add graphical representation of the conditions}
%$$
%\xymatrix{
%	x' \ar@{.>}[drr]^R \\
%	x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
%}
%\qquad\qquad
%\xymatrix{
%	&& z''  \\
%	x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
%}
%$$
%respectively, have already been studied
%in~\cite{plotkin:stirling:86}.
%
%In~\cite{simpson1994}, Simpson extends his basic sequent system for $\IK$ to the geometric axiom family. 
%%
%For example, you can add the following rule:
%$$\scalebox{.9}{$\vlinf{\boxbrn{4}}{\text{\footnotesize $\lb{u'}$ fresh}}{\B, \accs wv, \accs vu, \Left \SEQ \Right}{\B, \accs wv, \accs vu, \accs wu, \Left \SEQ \Right}$}$$
%to it and obtain a sound and complete system wrt.~$\IK$ plus the axiom
%$\ax{4}\colon \fm{(\DIA\DIA A \IMP \DIA A) \AND (\BOX A \IMP \BOX\BOX A)}$, that is, wrt.~to all frames in which $\rel$ is transitive.
%
%In~\cite{plotkin1986}, Plotkin and Stirling give a more general correspondence result than Theorem~\ref{thm:plotkin}, that is, for intuitionistic modal logic extended with a family of axioms wrt.~some classes of bi-relational frames.
%%
%For example, the frames that validate the axiom $\rn{4}_\rn\DIA \colon \fm{\DIA\DIA A \IMP \DIA A}$ are exactly the ones satisfying the condition:
%%\begin{center}
%($\diabrn{4}$) if $\accs wv$ and $\accs vu$, there exists a $\lb{u'}\in W$ s.t.~$\futs{u}{u'}$ and $\accs{w}{u'}$.
%%\end{center}
%
%Incorporating the preorder symbol into the syntax of our sequents allows us to also obtain a sound and complete proof system for the intuitionistic modal logic extended with axiom $\rn{4}_\rn\DIA$, by designing the following rule:
%$$\scalebox{.9}{$\vlinf{\diabrn{4}}{\text{\footnotesize $\lb{u'}$ fresh}}{\B, \accs wv, \accs vu, \Left \SEQ \Right}{\B, \accs wv, \accs vu, \futs{u}{u'}, \accs{w}{u'}, \Left \SEQ \Right}$}$$
%
%Consequently, we can easily translate the frame
%conditions in~\eqref{eq:4frame} into inference rules:
%\begin{equation}
%\label{eq:4rules}
%\vlinf{\rn{4^{\BOX}}}{\text{\footnotesize$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
%	\B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
%\qquad
%\vlinf{\rn{4^{\DIA}}}{\text{\footnotesize$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
%	\B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
%\end{equation}
%This allows us to define cut-free deductive systems for a wide range
%of logics that could not be treated before.
%
%%Therefore, we decompose further the formalism of labelled sequents and extend the reach of labelled deduction to the logics studied in~\cite{plotkin1986}.
%%%
%%These systems enjoy cut-elimination via usual arguments, the generality of the result is subject of ongoing study.
%
%
%\paragraph{Nested}
%Replace $\lrn\BOX$ and $\rrn\DIA$ above by the following:
%\[
%\vlinf{\lrn[t]\BOX}{}{\Cx{\lf{\BOX A}}}{\Cx{\lf{\BOX A}, \lf A}}
%\quad
%%\vlinf{\lrn[2]\BOX}{}{\Cx[1]{\lf{\BOX A}, \bBR{\BR{\Cx[2]}}}}{\Cx[1]{\bBR{\BR{\lf{\BOX A}, \Cx[2]}}}}
%\vlinf{\lrn[4]\BOX}{}{\Cx[1]{\lf{\BOX A}, \BR{\Cx[2]}}}{\Cx[1]{\lf{\BOX A}, \BR{\lf{\BOX A}, \Cx[2]}}}
%\quad
%\vlinf{\rrn[t]\DIA}{}{\Cx{\rt{\DIA A}}}{\Cx{\rt{\DIA A}, \rt A}}
%\quad
%\vlinf{\rrn[4]\DIA}{}{\Cx[1]{\rt{\DIA A}, \BR{\Cx[2]}}}{\Cx[1]{\rt{\DIA A}, \BR{\rt{\DIA A}, \Cx[2]}}}
%\]
%\[\mbox{\hbox to .9\linewidth{\dotfill}}\]
%\[
%\vlinf{\rn{refl_{\BR{\cdot}}}}{}{\Cx[1]{\Cx[2]}}{\Cx[1]{\BR{\Cx[2]}}}
%\qquad
%\vlinf{\rn{trans_{\BR{\cdot}}}}{}{\Cx[1]{\BR{\Cx[2], \BR{\Cx[3]}}}}{\Cx[1]{\BR{\Cx[2]}, \BR{\Cx[3]}}}
%\]

>>>>

This lead us to develop a \emph{fully structured} approach to intuitionistic modal proof theory capturing both the modal accesibility relation and the intuitionistic preorder relation. 
%
A fully labelled framework, described succintly in Section~\ref{sec:labelled}, has already allowed us to obtain modular systems for all intuitionistic Scott-Lemmon logics~\cite{marin:morales:strassburger:hal}. 
%
In an attempt to make this system amenable for proof-search and decision procedures, we have started investigated a fully nested framework, presented in Section~\ref{sec:nested}.
%
We would be particularly interested in a suitable system for logic $\mathsf{IS4}$ whose decidability is not known.

%We present a labelled sequent system and a nested sequent system for intuitionistic modal logics capturing both the modal accesibility relation and the intuitionistic preorder relation. 
%
%In the labelled calculus we made explicity both relations with $\rel$ and $\le$ into the system. 
%
%In the nested sequent system, we represent this relations with one bracket in a sequent (for the accesibility relation) and a with two brackets in a sequent (preorder relation).  
%

@@

We write $\lb x \Vdash \fm a$ if $\fm a \in V(\lb x)$ and, by definition, it is never the case that $\lb x \Vdash \fm{\BOT}$. %and always have $\lb w \Vdash \fm{\TOP}$
%
The relation $\Vdash$ is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:\vspace*{-.1cm}
\begin{align}
	&\lb x \Vdash \fm{A \AND B} && \text{iff}\qquad 
	\lb x \Vdash \fm A \text{ and } \lb x \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{A \OR B} && \text{iff}\qquad 
	\lb x \Vdash \fm A \text{ or } \lb x \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{A \IMP B} && \text{iff}\qquad 
	\text{for all } \lb y \text{ with } \futs xy\text{, if }\lb y \Vdash \fm A\text{ then }\lb y \Vdash \fm B
	\nonumber\\
	&\lb x \Vdash \fm{\BOX A} && \text{iff}\qquad
	\text{for all }\lb y\text{ and }\lb z\text{ with }\futs xy\text{ and }\accs yz, \lb z \Vdash \fm A %\hfill $(\ast)$
	\label{eq:box}\\
	&\lb x \Vdash \fm{\DIA A} && \text{iff}\qquad
	\text{there exists a }\lb y\text{ such that }\accs xy\text{ and }\lb y \Vdash \fm A
	\nonumber
\end{align}

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
		%		
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%
		\item\label{iii} $A$ is provable in $\labIKp$
		%
%		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}

@@

\begin{definition}
	A two-sided intuitionistic \emph{fully nested sequent} is constructed from the grammar:
	$\Cx \coloncolonequals \varnothing \mid \lf A, \Cx \mid \rt A, \Cx \mid \BR{\Cx} \mid \bBR{\Cx}$,
	%
%	\noindent
	where the $\lf{(\cdot)}$ and $\rt{(\cdot)}$ annotations are used to indicate that the formulas would occur on the left-hand-side or right-hand-side of a sequent respectively in the absence of the sequent arrow. 
\end{definition}

@@

@@

\section{Extensions: example of transitivity}

As mentionned in the introduction, one of our motivation is to investigate decision procedure for axiomatic extensions of $\IK$, for instance for $\mathsf{IS4}$, intuitionistic logic of reflexive transitive frames.
%
So, we will illustrate the benefit of our approach on transitivity here.

%\paragraph{Labelled}
%
%The frame condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
%inference rule
%\begin{equation}
%\label{eq:Rtrans}
%\vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
%	\B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
%\end{equation}
%where $\B$ stands for a set of relational atoms, and $\Gamma$ and
%$\Delta$ for multi-sets of labelled formulas.

The frame condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$) can be axiomatised by the conjunction of the two versions of the $\ax 4$-axiom:\vspace*{-.2cm}
%\begin{equation*}
%\label{eq:4ax}
$$
\ax{4_{\BOX}}\colon\fm{\BOX A\IMP \BOX\BOX A}
\qquad\qquad\qquad
\ax{4_{\DIA}}\colon\fm{\DIA\DIA A\IMP\DIA A}
$$%\end{equation*}
%
which are equivalent in classical modal logic. 
%
However, in intuitionistic modal logic they are not and they can be added independently to the logic $\IK$. %(an intuitionistic version of $\K$). 
%
From~\cite{plotkin1986} we know they are in correspondance with the following frame conditions:\vspace*{-.2cm}
\begin{equation}
\label{eq:4frame}
\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{u}.\:\futs x{u}\AND\accs {u}z)
\quad
\forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{u}.\:\futs z{u}\AND\accs {x}{u})
\end{equation}
%  \todo{add graphical representation of the conditions}
%$$
%\xymatrix{
%	u \ar@{.>}[drr]^R \\
%	x \ar@{.>}[u]^\le \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z
%}
%\qquad\qquad
%\xymatrix{
%	&& u  \\
%	x  \ar@{.>}[urr]^R \ar@{->}[r]_{R} & y \ar@{->}[r]_{R} & z \ar@{.>}[u]^\le
%}
%$$
%respectively%, have already been studied in~\cite{plotkin:stirling:86}.

Following Simpson~\cite{simpson1994} we could extend our basic sequent system for $\IK$ to $\mathsf{IK4} = \IK+(\ax{4_{\BOX}}\AND\ax{4_{\DIA}})$ with the rule:\vspace*{-.2cm}
%
%For example, you can add the following rule:
$$\small
\vlinf{\rn{4}}{}{\B, \accs wv, \accs vu, \Left \SEQ \Right}{\B, \accs wv, \accs vu, \accs wu, \Left \SEQ \Right}$$
%to it and obtain a sound and complete system wrt.~$\IK$ plus the axiom
%$\ax{4}\colon \fm{(\DIA\DIA A \IMP \DIA A) \AND (\BOX A \IMP \BOX\BOX A)}$, that is, wrt.~to all frames in which $\rel$ is transitive.

%In~\cite{plotkin1986}, Plotkin and Stirling give a more general correspondence result than Theorem~\ref{thm:plotkin}, that is, for intuitionistic modal logic extended with a family of axioms wrt.~some classes of bi-relational frames.
%%
%For example, the frames that validate the axiom $\rn{4}_\rn\DIA \colon \fm{\DIA\DIA A \IMP \DIA A}$ are exactly the ones satisfying the condition:
%%\begin{center}
%($\diabrn{4}$) if $\accs wv$ and $\accs vu$, there exists a $\lb{u'}\in W$ s.t.~$\futs{u}{u'}$ and $\accs{w}{u'}$.
%%\end{center}

Incorporating the preorder symbol into the syntax, however,  allowed us to %also obtain a sound and complete proof system for the intuitionistic modal logic extended with axiom $\rn{4}_\rn\DIA$, by designing the following rule:
%$$\scalebox{.9}{$\vlinf{\diabrn{4}}{\text{\footnotesize $\lb{u'}$ fresh}}{\B, \accs wv, \accs vu, \Left \SEQ \Right}{\B, \accs wv, \accs vu, \futs{u}{u'}, \accs{w}{u'}, \Left \SEQ \Right}$}$$
%Consequently, we can easily 
translate the conditions in~\eqref{eq:4frame} into independent inference rules for $\ax{4_{\BOX}}$ and $\ax{4_{\DIA}}$:\vspace*{-.2cm}
\begin{equation*}\small
\label{eq:4rules}
\vlinf{\rn{4_{\BOX}}}{\text{\footnotesize$\lb{u}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {u}z, \futs x{u},\Left \SEQ \Right}
\quad
\vlinf{\rn{4_{\DIA}}}{\text{\footnotesize$\lb{u}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x}{u}, \futs z{u},\Left \SEQ \Right}  
\end{equation*}
%This allows us to define cut-free deductive systems for a wide range of logics that could not be treated before.

%Therefore, we decompose further the formalism of labelled sequents and extend the reach of labelled deduction to the logics studied in~\cite{plotkin1986}.
%%
%These systems enjoy cut-elimination via usual arguments, the generality of the result is subject of ongoing study.

A great interest of a nested system is that the rules for axiomatic extensions are usually more suitable for proof-search procedures that in the labelled setting.
%
Previous nested systems for intuitionistic modal logics~\cite{strassburger2013,kuznets:strassburger:maehara} can be extended from $\IK$ to $\mathsf{IK4}$ by simply adding the following rules:\vspace*{-.2cm}
\[\small
%\vlinf{\lrn[t]\BOX}{}{\Cx{\lf{\BOX A}}}{\Cx{\lf{\BOX A}, \lf A}}
%\quad
%\vlinf{\lrn[2]\BOX}{}{\Cx[1]{\lf{\BOX A}, \bBR{\BR{\Cx[2]}}}}{\Cx[1]{\bBR{\BR{\lf{\BOX A}, \Cx[2]}}}}
\vlinf{\lrn[4]\BOX}{}{\Cx[1]{\lf{\BOX A}, \BR{\Cx[2]}}}{\Cx[1]{\lf{\BOX A}, \BR{\lf{\BOX A}, \Cx[2]}}}
%\quad
%\vlinf{\rrn[t]\DIA}{}{\Cx{\rt{\DIA A}}}{\Cx{\rt{\DIA A}, \rt A}}
\quad
\vlinf{\rrn[4]\DIA}{}{\Cx[1]{\rt{\DIA A}, \BR{\Cx[2]}}}{\Cx[1]{\rt{\DIA A}, \BR{\rt{\DIA A}, \Cx[2]}}}
\]

The generalisation of these results to our fully structured framework is subject of ongoing study.
